---
title: 'STA304'
author: "Neil Montgomery"
date: "2016-03-03"
output: 
  ioslides_presentation: 
    css: 'styles.css' 
    widescreen: true 
    transition: 0.001
---
\newcommand{\E}[1]{E{\left(#1\right)}}
\newcommand{\flist}[2]{\{#1_1, #1_2, \ldots, #1_#2\}}
\newcommand{\fulist}[3]{\{#1_{{#2}1}, #1_{{#2}2}, \ldots, #1_{{#2}{#3}}\}}
\renewcommand{\bar}[1]{\overline{#1}}
\newcommand{\SE}[1]{\sqrt{\hat{V}(#1)}}

# ratio, regression, difference estimation

## overview { .build }

In most actual samples, more than one measurement is available on each unit sampled.

Let's consider the case of two measurements, generically called $y$ and $x$. 

The population is now $\{(y_1, x_1),\ldots,(y_N, x_N)\}$ and the sample is $\{(y_1, x_1),\ldots,(y_n, x_n)\}$ (with the usual abuse of notation.)

Sometimes the population parameter of interest is still the total or mean of the $y$. Let's call these numbers $\tau_y$ and $\mu_y$. We'll also make use of $\tau_x$ and $\mu_x$.

If the $y_i$ and the $x_i$ in the population are *correlated*, it is possible to get improved estimates of $\tau_y$ and $\mu_y$.

Also, sometimes *the actual quantity of interest* is the population ratio itself $R = \tau_y/\tau_x = \mu_y/\mu_x$.

## a note, and some examples { .build }

Note: stratified sampling was a way to use additional information about the units *to design the sampling itself in a different way that can be better*. This part of the course concerns using additional information *in the estimation procedures*.

Examples:

* Consumer Price Index (CPI): a "basket" of consumer goods has its prices measured each month. The $y$ variable is this month's price and the $x$ variable is the previous month's price. The ratio of the total basket price is the proportion of increase/decrease in prices.

* "The average amount of screen time per child in households." Unit: household. $y$: total screen time. $x$: number of children. 

* (Section 6.2): Total amount of sugar in a truckload of oranges. $y$ is sugar per orange and $x$ is the weight of an orange.

## estimating the population ratio { .smaller }

I'll use the salary data (a sample from a population of $N$ = 750) from question 6.10 in the textbook as an example to motivate the concepts. 

```{r, echo=FALSE, message=FALSE}
library(rio)
library(dplyr)
library(knitr)
salary <- import("Ch 6 Exercises.xls", sheet="6.10")
kable(salary)
N <- 750
```

## estimating the population ratio { .build }

$x$ is the *Past* salary and $y$ is the *Present*. The goal is to estimate the rate of salary increase for the population. This is the ratio:
$$R = \frac{\tau_y}{\tau_x} = \frac{\mu_y}{\mu_x}$$

The obvious way to estimate $R$ is to use the ratio of the sample totals or sample means:
$$\hat R = \frac{\hat\tau_y}{\hat\tau_x} = \frac{\sum_{i=1}^n y_i}{\sum_{i=1}^n x_i} = \frac{\bar y}{\bar x} = \frac{\hat\mu_y}{\hat\mu_x}$$
which the book mystifyingly calls $r$ rather than $\hat R$, so I'll respect that notation.

The basic challenge is that the denominator is random. 

## variance of $r$ - I{ .build }

It turns out $r$ is slightly biased for $R$, but mainly with very small sample sizes. **So we'll go with the assumption that the sample size is not very small**, giving $\E r \approx R$. 

The end goal is to get $V(r) \approx E\left[(r - R)^2\right]$. Start with:

$$r - R = \frac{\bar y}{\bar x} - R = \frac{\bar y - R \bar x}{\bar x}
\approx \frac{\bar y - R \bar x}{\mu_x}$$

Then:

$$V(r) \approx E\left[\left(\frac{\bar y - R \bar x}{\mu_x}\right)^2\right] = \frac{1}{\mu_x^2}E\left[\left(\bar y - R \bar x\right)^2\right]$$

## variance of $r$ - II { .build }

$\left(\bar y - R \bar x\right)$ is actually a pretty simple object. Let:

$$r_i = y_i - R x_i$$

Then $\{r_1,r_2,\ldots,r_n\}$ is a SRS and $\bar r = \left(\bar y - R \bar x\right)$ is a sample average with mean 0 and variance that can be copied from the basic SRS theory:
$$V(\bar r) = \frac{s^2_r}{n}\left(1-\frac{n}{N}\right)$$
with
$$s^2_R = \frac{\sum_{i=1}^n \left(r_i - \bar r\right)^2}{n-1}$$

## variance of $r$ - III { .build }

Putting it all together and we get:

$$V(r) \approx \frac{1}{\mu_x^2}\frac{s^2_R}{n}\left(1-\frac{n}{N}\right)$$

There are a few unknowns in there. $R$ is never known, and $\mu_x$ may or may not be known. The conclusion is:

$$\hat V(r) \approx \frac{1}{\mu_x^2}\frac{s^2_r}{n}\left(1-\frac{n}{N}\right)
$$

where $s^2_r$ is $s^2_R$ with $r$ used in place of $R$, and use $\bar x$ if $\mu_x$ is unknown. 

## salary example

Let's estimate the population salary rate of increase and place the usual bound on the error of estimation $\left(B = 2\sqrt{\hat V(r)}\right)$. 

```{r, echo=FALSE, fig.height=3, fig.width=3, fig.align='center'}
salary %>% 
  summarize(x_bar = mean(Past), y_bar = mean(Present), 
            r = y_bar/x_bar, "s^2_r" = var(Present - r*Past),
            B = 2*sqrt((1-nrow(salary)/N)/x_bar^2*`s^2_r`/nrow(salary))) -> salary_summ
kable(salary_summ)
```

&nbsp;

```{r, echo=FALSE, fig.height=3, fig.width=3, fig.align='center'}
library(ggplot2)
salary %>% 
  ggplot(aes(x=Past, y=Present)) + geom_point()
```

## using $x$ and ratio technique for estimation of $\tau_y$ { .build }

Ratio estimation might make estimating $\tau_y$ *better*, or it might even be the only feasible way to do it at all, if $N$ isn't known and is too difficult to determine, but $\tau_x$ *is* known.

An example of the latter case the total amount of sugar $\tau_y$ in a truck of oranges example. (Section 6.2 and Example 6.2). Here $\tau_x$ is the total weight of the oranges.

In general the usual estimator of $\tau_y$ is $N\bar y$, but always requires $N$. Another option is:
$$\hat\tau_y = r\tau_x$$
If the $y_i$ and $x_i$ are correlated then this is a better estimator, and if $N$ is unknown this can be the only feasible estimator.

## estimated variance of $\hat\tau_y$

Option 1:
$$\hat V(\hat\tau_y) = N^2\left(1-\frac{n}{N}\right)\frac{s^2_r}{n}$$

Option 2 ($N$ unknown)::
$$\hat V(\hat\tau_y) = (\tau_x^2)\frac{1}{\mu_x^2}\frac{s^2_r}{n}\left(1-\frac{n}{N}\right)$$
with $\bar x$ for $\mu_x$ if required.

## sugar in oranges example { .build }

(Raw data not available). A sample of $n=10$ oranges was taken and the sugar content $y_i$ and weight $x_i$ measured. The total weight $\tau_x$ of all oranges was 1800 pounds. 

The estimate of the total sugar in pounds is:

$$\hat\tau_y = r\tau_x = \frac{\sum y_i}{\sum x_i} = \frac{0.246}{4.35}(1800) = 101.79$$

The bound $B = 2\sqrt{\hat V(\hat\tau_y)}$ requires also to know that $s^2 = (0.0024)^2$. Plugging all into the formula gives a bound of 6.3
